---
title: 440-individual-project
authors:
  - name: Louis Choo-Choy
    affiliation: Duke University
    roles: writing
    corresponding: true
bibliography: references.bib
---

## Packages

```{r}
library(tidyverse)
```

## Exploratory Data Analysis

```{r}

```

## Introduction

We are interested in the relationship between perceived satisfaction with quality of life and length of time living in the Durham. This could be used as a measure of whether the living standards of Durham have improved over time. As secondary questions, we are interested in whether a resident being a renter or an owner affects this perceived satisfaction and whether the satisfaction of people living in the same neighborhood is independent of one another.\

Variables of relevance:

Q1. Satisfaction with Major Categories of Services Provided by the City and County

Q3. Level of Satisfaction with Items That Influence Perceptions of Durham

Q27. Approximately how many years have you lived in Durham?

Q32. Do you own or rent your current residence?

Zip

PAC (Partners Against Crime)

```{r}

resident_survey <- read_csv("data/resident_survey.csv")
```
